---
title: "Vis 1"
output: github_document
---

```{r setup, include = FALSE}
library(tidyverse)
library(ggridges)
```

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())
```

Making our first plot :) 

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) + 
   geom_point()

```

```{r}
weather_df |>
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point()
```

Check why some rows are missing. 

```{r}
ggp_weather = 
  weather_df |>
  ggplot(aes(x = tmin, y = tmax)) 

ggp_weather + geom_point()

```


##Fancier scatterplots!

```{r}
weather_df |>
ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .3, size = .8) + 
  geom_smooth(se = FALSE)
```

Where you define aesthetics can matter

```{r}
weather_df |>
ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .3, size = .8) + 
  geom_smooth(se = FALSE)
```

Use faceting real quick

```{r}
weather_df |>
ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .3) +
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name)
```

Let's make a somewhat more interesting scatterplot
```{r}
weather_df |>
ggplot(aes(x = date, y = tmax, color = name)) + 
  geom_point(aes(size = prcp), alpha = .3) +
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name)
```

##Learning assessment plot

Write a code chain that starts with weather_df; focuses only on Central Park, converts temperatures to Fahrenheit, makes a scatterplot of min vs. max temperature, and overlays a linear regression line (using options in geom_smooth())

```{r}
weather_df |>
filter(name == "CentralPark_NY") |> 
  mutate(
    tmax_fahr = tmax * (9 / 5) + 32,
    tmin_fahr = tmin * (9 / 5) + 32) |> 
  ggplot(aes(x = tmin_fahr, y = tmax_fahr)) +
  geom_point(alpha = .5) + 
  geom_smooth(method = "lm", se = FALSE)
```


