---
title: "Vis 2"
output: github_document
---

```{r setup, include = FALSE}
library(tidyverse)
library(ggridges)
```

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())
```

Making our first plot :) 

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) + 
   geom_point()

```

```{r}
weather_df |>
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point()
```

Check why some rows are missing. 

```{r}
ggp_weather = 
  weather_df |>
  ggplot(aes(x = tmin, y = tmax)) 

ggp_weather + geom_point()

```


##Fancier scatterplots!

```{r}
weather_df |>
ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .3, size = .8) + 
  geom_smooth(se = FALSE)
```

Where you define aesthetics can matter

```{r}
weather_df |>
ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .3, size = .8) + 
  geom_smooth(se = FALSE)
```

Use faceting real quick

```{r}
weather_df |>
ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .3) +
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name)
```

Let's make a somewhat more interesting scatterplot
```{r}
weather_df |>
ggplot(aes(x = date, y = tmax, color = name)) + 
  geom_point(aes(size = prcp), alpha = .3) +
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name)
```

##Learning assessment plot

Write a code chain that starts with weather_df; focuses only on Central Park, converts temperatures to Fahrenheit, makes a scatterplot of min vs. max temperature, and overlays a linear regression line (using options in geom_smooth())

```{r}
weather_df |>
filter(name == "CentralPark_NY") |> 
  mutate(
    tmax_fahr = tmax * (9 / 5) + 32,
    tmin_fahr = tmin * (9 / 5) + 32) |> 
  ggplot(aes(x = tmin_fahr, y = tmax_fahr)) +
  geom_point(alpha = .5) + 
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
weather_df |>
ggplot(aes(x = date, y = tmax, color = name)) + 
  geom_smooth(se = FALSE) 

```

```{r}
weather_df |>
ggplot(aes(x = tmax, y = tmin)) + 
  geom_hex()
```

```{r}
weather_df |>
ggplot(weather_df) + geom_point(aes(x = tmax, y = tmin), color = "blue")
ggplot(weather_df) + geom_point(aes(x = tmax, y = tmin, color = "blue"))

```

##Univariate plots

```{r}
weather_df |>
  ggplot(aes(x = tmax, fill = name)) + 
  geom_histogram()
```

how would i fix this? maybe facet?
```{r}
weather_df |>
  ggplot(aes(x = tmax, fill = name)) + 
  geom_histogram(position = "dodge", binwidth = 2) +
  facet_grid(. ~ name)
```


```{r}
weather_df |>
  ggplot(aes(x = name, y = tmin, fill = name)) +
  geom_boxplot()
```

violin plots
```{r}
weather_df |>
ggplot(aes(x = name, y = tmax, fill = name)) + 
  geom_violin() 
  
```

ridge plot

```{r}
weather_df |>
  ggplot(aes(x = tmax, y = name)) + 
  geom_density_ridges(scale = .85)
```

##Learning Assessment

Make plots that compare precipitation across locations. Try a histogram, a density plot, a boxplot, a violin plot, and a ridgeplot; use aesthetic mappings to make your figure readable

First a density plot
```{r}
weather_df |>
ggplot(aes(x = prcp)) + 
  geom_density(aes(fill = name), alpha = .5) 
```

Next a ridge plot:
```{r}
ggplot(weather_df, aes(x = prcp, y = name)) + 
  geom_density_ridges(scale = .85)
```

Last a boxplot:
```{r}
ggplot(weather_df, aes(y = prcp, x = name)) + 
  geom_boxplot()
```

```{r}
weather_df |> 
  filter(prcp > 10, prcp < 1000) |> 
  ggplot(aes(x = prcp, fill = name)) + 
  geom_density(alpha = .3) 
```

##Saving and embedding plots

```{r}
ggp_weather = 
  weather_df |>
  ggplot(aes(x = date, y = tmax, color = name)) + 
  geom_point() 

ggsave("ggp_weather.pdf", ggp_weather, width = 8, height = 6)

```

embedding plots 

```{r, fig.width = 6, fig.height = 12}
weather_df |>
  ggplot(aes(x = date, y = tmax, color = name)) + 
  geom_point() 
```


Make a scatterplot but fancy this time.

```{r}
weather_df |>
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .3)
  labs(
    title = "Temperatre scatterplot",
    x = "minimum temp (C)",
    y = "maximum temp (C)",
    color = "Location",
    caption = "Weather data taken from rnoaa package for three stations."
  )
```

Scales -- start with `x` and `y` and then do `color`

```{r}
weather_df |>
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .3)
  labs(
    title = "Temperatre scatterplot",
    x = "minimum temp (C)",
    y = "maximum temp (C)",
    color = "Location",
    caption = "Weather data taken from rnoaa package for three stations."
  ) +
    scale_x_continuous(
      breaks = c(-15, 0, 20),
      labels = c("-15C", "0", "20")
    ) +
    scale_y_continuous(
      limits = c(0, 30),
      transform ="sqrt"
    )
```

Look at color: 


```{r}
weather_df |>
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .3) +
  labs(
    title = "Temperatre scatterplot",
    x = "Minimum Temp (C)",
    y = "Maximum Temp (C)",
    color = "Location",
    caption = "Weather data taken from rnoaa package for three stations."
  ) + 
    viridis::scale_color_viridis(discrete = TRUE)
```

### Themes







